Dependent Object Types

### 2012-08-05 article.tex
some counterexs from report

% Sugar
\newcommand{\arrow}[2]{#1\rightarrow_s#2}
\newcommand{\fun}[4]{\textbf{fun } (#1:#2)\;#3\;#4}
\newcommand{\app}[2]{(\textbf{app }#1\;#2)}
\newcommand{\cast}[2]{(\textbf{cast }#1\;#2)}

@@

@@

>>>>

% Sugar
\newcommand{\arrow}[2]{#1\rightarrow_s#2}
\newcommand{\fun}[4]{\textbf{fun } (#1:#2)\;#3\;#4}
\newcommand{\app}[2]{(\textbf{app }#1\;#2)}
\newcommand{\mlapp}[2]{(\textbf{app }#1\;\\&#2)}
\newcommand{\cast}[2]{(\textbf{cast }#1\;#2)}

@@

We prove that the DOT calculus is type-safe using logical relations in
section~\ref{dot-type-safety}. However, we first tried to prove the
calculus type-safe using the standard theorems of preservation and
progress. Unfortunately, for the calculus as presented, and any of its
variants that we devised, preservation doesn't hold. In this section,
we review some of the most salient counterexamples to preservation
that we found. These counterexamples have been checked with PLT Redex.

\subsection{\texorpdfstring{\textsc{term-$\ni$}}{Term-Mem} Restriction}\label{term_mem}

There are two membership ($t \ni D$) rules: one for when the term $t$
is a path, and one for an arbitrary term $t$. For paths, we can
substitute the self-references in the declarations, but we cannot do
so for arbitrary terms as the resulting types wouldn't be well-formed
syntactically. Hence, the $\textsc{term-$\ni$}$ has the restriction
that self-occurrences are not allowed. Here is a counterexample related
to this restriction.

Let $X$ be a shorthand for the type:
\begin{align*}
\Top & \mlrefine z {\\
&\ \Ldecl {L_a} \Top \Top\\
&\ \ldecl l {z.L_a}\\
}&
\end{align*}

Let $Y$ be a shorthand for the type:
\begin{align*}
\Top & \mlrefine z {\\
&\ \ldecl l \Top\\
}&
\end{align*}

Now, consider the term
\begin{align*}
&\mlnew u {X \ldefs{ l = u }}{
(\app {(\abs y {\Top \tfun Y} {\app y u})} {(\abs d \Top {(\app {(\abs x X x)} u)})}).l
}
\end{align*}

The term type-checks because the term $t=(\app {(\abs y {\Top \tfun Y}
  {\app y u})} {(\abs d \Top {(\app {(\abs x X x)} u)})})$ has type
$Y$, so we can apply $\textsc{term-$\ni$}$ for $l$. However, the term
$t$ eventually steps to $(\app {(\abs x X x)} u)$ which has type $X$,
so we cannot apply $\textsc{term-$\ni$}$ for $l$ because of the
self-reference ($z.L_a$).

\subsection{Expansion Lost}\label{narrowing_exp}

Expansion is not preserved by narrowing. Here, we create two type
selections that are mutually recursive in their upper bounds after
narrowing: $z_0.C_2$ initially expands, but after narrowing, $z_0.C_2$
expands to what $z_0.A_2$ expands to, which expands to what $z_0.A_1$
expands to, which expands to what $z_0.A_2$ expands to, and thus we
have an infinite expansion. Thus, the last new expression initially
type-checks, but after narrowing, it doesn't because the precise
expansion needed by $\textsc{new}$ cannot be inferred.

\begin{align*}
&\mlnew {x_0} {\Top \mlrefine z { \Ldecl {A_1} \Bot {\Top \mlrefine z {\\
&\gap\Ldecl {A_2} \Bot \Top\\
&\gap\Ldecl {A_3} \Bot \Top\\
&\gap\Ldecl {C_2} \Bot {z.A_2}}}}\ldefs{}}{
\mlnew {x_1} {\Top \refine z {\Ldecl {C_1} \Bot {\Top \refine z {\Ldecl {A_1} \Bot {x_0.A_1}}}}\ldefs{}}}{
\mlnew {x_2} {{x_1.C_1} \refine z {\Ldecl {A_1} \Bot {{x_0.A_1} \refine z {\Ldecl {A_2} \Bot {z.A_3}}}}\ldefs{}}}{
\mlnew {x_3} {{x_1.C_1} \refine z {\Ldecl {A_1} \Bot {{x_0.A_1} \refine z {\Ldecl {A_3} \Bot {z.A_2}}}}\ldefs{}}}{
\mlapp {\abs x {x_1.C_1} {(\abs {z_0} {x.A_1 \tand x_3.A_1} {\\&\gap\gap\gap\new z {z_0.C_2} {\app {(\abs x \Top x)} z}})}} {\ x_2}}
\end{align*}

\subsection{Well-Formedness Lost}\label{narrowing_wf}

Even well-formedness is not preserved by narrowing. The trick is that
if the lower bound of a type selection is not $\Bot$, then the
bounds needs to be checked for well-formedness. Here, we create two
type selections that are mutually recursive in their bounds
after narrowing. $y.A$ is initially well-formed, but after narrowing,
it isn't because we run into an infinite derivation trying to prove
the well-formedness of its bounds.

\begin{align*}
&\mlnew v {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B {z.A} {z.A}}}} \ldefs{}}{
\mlapp {\abs x {\Top \refine z {\Ldecl L \Bot {\Top \refine z {\Ldecl A \Bot \Top, \Ldecl B \Bot \Top}}}} {\\&\gap
\mlnew z {\Top \refine z {\ldecl l {\Bot \tfun \Top}}\mlldefs{\\&\gap\gap l = \abs y {x.L \tand {\Top \refine z {\Ldecl A {z.B} {z.B}, \Ldecl B \Bot \Top}}} {\\&\gap\gap\gap\abs a {y.A} {\app {(\abs x \Top x)} a}}}}{
\ \app {(\abs x \Top x)} z
}}}{\ v}}
\end{align*}

\subsection{Path Equality}\label{patheq}

We need to be able to relate path-dependent types after reduction. The
original DOT calculus doesn't have any rules for dealing with
path-equality. Here is an example which motivates the need for
path-equality provisions.

\begin{align*}
&\mlnew b {\Top \mlrefine z {&&\Ldecl X \Top \Top&\\
&&&\ \ldecl l {z.X}&}\ldefs{l = b}}{
\mlnew a {\Top \mlrefine z {\ldecl i {\Top \mlrefine z {&&&\\
&&&\Ldecl X \Bot \Top\\
&&&\ldecl l {z.X}}}&}\ldefs{i = b}}}{
\app {(\abs x \Top x)} {(\app {(\abs x {a.i.X} x)} {a.i.l})}}
\end{align*}

$a.i.l$ reduces to $b.l$. $b.l$ has type $b.X$, so we need $b.X <:
a.i.X$. This cannot be established with the current rules: it is not
true in general, but true here because $a.i$ reduces to $b$. Hence,
the need for acknowledging path equality.

@@

\section{Discussion}\label{discussion}
